---
title: "AE 08 - The Office - Predicting IMDB Ratings"
author: "Kaden Zhu"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Introduction

In this application exercise, you'll build a predictive model for IMDB ratings of episodes of *The Office* using the **tidymodels** framework. You'll practice feature engineering, creating modeling workflows, and evaluating model performance.

**In this application exercise, you will:**

- Engineer features from text data (episode scripts)
- Create indicator variables for special episodes
- Build a modeling dataset
- Split data into training and testing sets
- Specify and fit linear regression models
- Create preprocessing recipes
- Build modeling workflows
- Perform cross-validation
- Compare model performance

**Estimated time:** 60-90 minutes

---

## Getting Started

### Fork and Clone Workflow

**Quick reminder:** Work in YOUR forked `application-exercises` repository.

1. Navigate to the `ae-08-office` folder in YOUR fork
2. Open `ae-08-office.Rmd`
3. Update the YAML with your name
4. Knit to verify it works

---

## Packages
```{r load-packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(tidymodels)
library(schrute)     # Contains The Office data
library(lubridate)   # For working with dates
```

**What these packages do:**

- **tidyverse:** Data wrangling and visualization
- **tidymodels:** Modeling framework (replaces caret)
- **schrute:** Contains transcripts from The Office
- **lubridate:** Working with dates

---

## The Data

Use `theoffice` data from the [**schrute**](https://bradlindblad.github.io/schrute/) package to predict IMDB scores for episodes of The Office.
```{r explore-data}
glimpse(theoffice)
```

**Understanding the data:**

- Each row is a line of dialogue
- Multiple rows per episode (one per line)
- Variables include: season, episode, character, text, imdb_rating, total_votes, air_date

**Fix `air_date` for later use:**
```{r fix-air-date}
theoffice <- theoffice %>%
  mutate(air_date = ymd(as.character(air_date)))
```

**What this does:**

- Converts `air_date` to proper Date format
- `ymd()`: Parses "year-month-day" format

---

## Our Goal

We will:

- Engineer features based on episode scripts
- Train a model
- Perform cross validation
- Make predictions

**Note:** The episodes listed in `theoffice` don't match the ones listed in the data we used in the cross validation lesson.
```{r check-episodes}
theoffice %>%
  distinct(season, episode)
```

---

## Exercise 1: Calculate Line Percentages

Calculate the percentage of lines spoken by Jim, Pam, Michael, and Dwight for each episode of The Office.

**Why?** These are main characters, and the amount they speak might relate to episode ratings.

**Task 1.1:** Create a dataset with line percentages.
```{r lines, eval=TRUE}
office_lines <- theoffice %>%
  group_by(season, episode) %>%
  summarise(
    total_lines = n(),
    jim_lines = sum(character == "Jim"),
    pam_lines = sum(character == "Pam"),
    michael_lines = sum(character == "Michael"),
    dwight_lines = sum(character == "Dwight"),
    .groups = "drop"
  ) %>%
  mutate(
    jim_pct = jim_lines / total_lines,
    pam_pct = pam_lines / total_lines,
    michael_pct = michael_lines / total_lines,
    dwight_pct = dwight_lines / total_lines
  )

office_lines
```

**Hints:**

- Group by `season` and `episode`
- Character names: "Jim", "Pam", "Michael", "Dwight"
- `sum(character == "Jim")` counts lines spoken by Jim
- `n()` gives total number of rows (lines) in each group

**Task 1.2:** Which episode has the highest percentage of Michael lines?
```{r highest-michael, eval=TRUE}
office_lines %>%
  arrange(desc(michael_pct)) %>%
  select(season, episode, michael_pct)
```

**Your answer:** Season 1, Episode 5 (45.2 percent of lines)

---

## Exercise 2: Identify Special Episodes

Identify episodes that touch on Halloween, Valentine's Day, and Christmas.

**Why?** Special holiday episodes might have different ratings patterns.

**Task 2.1:** Search episode text for holiday keywords.
```{r special-episodes, eval=TRUE}
office_episodes <- theoffice %>%
  group_by(season, episode) %>%
  summarise(
    halloween = sum(str_detect(str_to_lower(text), "halloween")) > 0,
    valentine = sum(str_detect(str_to_lower(text), "valentine")) > 0,
    christmas = sum(str_detect(str_to_lower(text), "christmas")) > 0,
    .groups = "drop"
  ) %>%
  mutate(
    halloween = as.numeric(halloween),
    valentine = as.numeric(valentine),
    christmas = as.numeric(christmas)
  )

office_episodes
```

**Hints:**

- Search for keywords: "halloween", "valentine", "christmas"
- `str_detect()`: Returns TRUE if pattern found
- `str_to_lower()`: Converts text to lowercase
- `sum(...) > 0`: TRUE if found at least once
- `as.numeric(TRUE)` = 1, `as.numeric(FALSE)` = 0

**Task 2.2:** How many episodes mention each holiday?
```{r count-holidays, eval=TRUE}
office_episodes %>%
  summarise(
    halloween_episodes = sum(halloween),
    valentine_episodes = sum(valentine),
    christmas_episodes = sum(christmas)
  )
```

**Your answers:**

Halloween episodes: 10
Valentine episodes: 7
Christmas episodes: 34

---

## Exercise 3: Create Modeling Dataset

Put together a modeling dataset that includes features you've engineered. Also add an indicator variable called `michael` which takes the value `1` if Michael Scott (Steve Carrell) was there, and `0` if not.

**Note:** Michael Scott left the show at the end of Season 7.

**Task 3.1:** Get one row per episode with all features.
```{r office-df, eval=TRUE}
office_info <- theoffice %>%
  group_by(season, episode) %>%
  summarise(
    episode_name = first(episode_name),
    imdb_rating = first(imdb_rating),
    total_votes = first(total_votes),
    air_date = first(air_date),
    .groups = "drop"
  )

office_df <- office_info %>%
  left_join(office_lines, by = c("season", "episode")) %>%
  left_join(office_episodes, by = c("season", "episode")) %>%
  mutate(michael = if_else(season <= 7, 1, 0))

office_df
```

**Hints:**

- Join with `office_lines`
- Join with `office_episodes`
- Join by `"season"` and `"episode"`
- Michael left after season 7: `season <= 7`

**Task 3.2:** Verify the dataset.
```{r verify-office-df, eval=TRUE}
dim(office_df)

office_df %>%
  summarise(across(everything(), ~sum(is.na(.))))

office_df %>%
  count(season, michael)
```

**How many episodes are in the dataset?** 186

**Do seasons 1-7 have michael = 1?** Yes. Seasons 1 through 7 all show michael equal to 1. Seasons 8 and 9 show michael equal to 0.

---

## Exercise 4: Split the Data

Split the data into training (75%) and testing (25%).
```{r split, eval=TRUE}
set.seed(1122)

office_split <- initial_split(office_df, prop = 0.75)
office_train <- training(office_split)
office_test <- testing(office_split)

nrow(office_train)
nrow(office_test)
```

**Hints:**

- `initial_split(office_df, prop = 0.75)`
- `training(office_split)`
- `testing(office_split)`

**Training set size:** 139 episodes

**Testing set size:** 47 episodes

---

## Exercise 5: Specify a Model

Specify a linear regression model.
```{r model, eval=TRUE}
office_mod <- linear_reg() %>%
  set_engine("lm")
```

**Hint:** Use `"lm"` as the engine for linear regression.

---

## Exercise 6: Create a Recipe

Create a recipe that:

- Updates the role of `episode_name` to not be a predictor
- Removes `air_date` as a predictor  
- Uses `season` as a factor
- Removes all zero variance predictors
```{r recipe, eval=TRUE}
office_rec <- recipe(imdb_rating ~ ., data = office_train) %>%
  update_role(episode_name, new_role = "ID") %>%
  step_rm(air_date) %>%
  step_mutate(season = as.factor(season)) %>%
  step_zv(all_predictors())

office_rec
```

**Hints:**

- Formula: `imdb_rating ~ .` (predict rating with all variables)
- Data: `office_train`
- Update role of `episode_name` to `"ID"`
- Remove `air_date`
- Convert `season` to factor
- Remove zero variance from `all_predictors()`

---

## Exercise 7: Build a Workflow

Build a workflow for fitting the model specified earlier and using the recipe you developed to preprocess the data.
```{r workflow, eval=TRUE}
office_wflow <- workflow() %>%
  add_model(office_mod) %>%
  add_recipe(office_rec)

office_wflow
```

**Hints:**

- Add the model: `office_mod`
- Add the recipe: `office_rec`

---

## Exercise 8: Fit the Model

Fit the model to training data and interpret a couple of the slope coefficients.
```{r fit, eval=TRUE}
office_fit <- office_wflow %>%
  fit(data = office_train)

tidy(office_fit)
```

**Hint:** Fit to `office_train`

**Task 8.1:** What is the coefficient for `michael`?

**Your answer:** NA (rank deficient). Michael is a perfect linear function of the season factor dummies. Michael equals 1 for seasons 1 through 7 and 0 for seasons 8 and 9, so once the season factor enters the model the michael indicator carries no extra information and lm drops it. The Michael effect is absorbed into the season1 baseline plus the season8 and season9 dummies. Season 8 has coefficient 0.0915 and season 9 has coefficient 0.426, both small, while seasons 2 through 7 sit between 0.40 and 0.73 above season 1.

**Interpretation:**

Episodes with Michael Scott have an IMDB rating that is roughly 0.3 points higher than episodes without him on average, holding all else constant. You see this by comparing the average season dummy for the Michael seasons (around 0.6 above season 1) with the average season dummy for the non-Michael seasons (0.0915 and 0.426, averaging about 0.26 above season 1). The signed direction is higher.

**Task 8.2:** What is the coefficient for `jim_pct`?

**Your answer:** 0.862

**Interpretation:**

For every 1 unit increase in the proportion of lines spoken by Jim, the IMDB rating increases by 0.862 points on average, holding all else constant. Because jim_pct ranges from 0 to about 0.30, a more useful read is that a 10 percentage point rise in Jim's share of lines lifts the predicted rating by about 0.086 points. The p value of 0.728 shows this estimate is not statistically significant.

---

## Exercise 9: Cross-Validation

Perform 5-fold cross validation and view model performance metrics.
```{r cv, message=FALSE, eval=TRUE}
set.seed(345)
folds <- vfold_cv(office_train, v = 5)
folds

set.seed(456)
office_fit_rs <- office_wflow %>%
  fit_resamples(folds)

collect_metrics(office_fit_rs)
```

**Hints:**

- Create folds from `office_train`
- Use `v = 5` for 5-fold CV
- Fit `office_wflow` to `folds`
- Collect metrics from `office_fit_rs`

**Task 9.1:** What is the mean RMSE from cross-validation?

**Your answer:** 0.357

**Task 9.2:** What is the mean R-squared from cross-validation?

**Your answer:** 0.597

**Task 9.3:** What do these metrics tell you about model performance?

**Your answer:**

The model explains about 60 percent of the variance in IMDB ratings across held out folds. RMSE of 0.357 means predictions land within roughly 0.36 of a rating point on average, which is small relative to the 6.8 to 9.7 spread of ratings in the data. Performance is moderate. The features capture real signal beyond noise, but 40 percent of variance is unexplained, so structural drivers like writing quality, guest stars, and finale status are still missing from the feature set.

---

## Exercise 10: Compare Models

Use your model to make predictions for the testing data and calculate the RMSE. Also use the model developed in the cross validation lesson to make predictions for the testing data and calculate the RMSE as well. Which model did a better job in predicting IMDB scores for the testing data?

### New Model
```{r new-model, message=FALSE, eval=TRUE}
office_pred <- predict(office_fit, office_test) %>%
  bind_cols(office_test)

office_pred %>%
  rmse(truth = imdb_rating, estimate = .pred)
```

**Hints:**

- Predict using `office_fit` on `office_test`
- Bind predictions to `office_test`
- RMSE: truth = `imdb_rating`, estimate = `.pred`

**New model RMSE:** 0.400

### Old Model
```{r old-model, eval=TRUE}
office_mod_old <- linear_reg() %>%
  set_engine("lm")

office_rec_old <- recipe(imdb_rating ~ season + episode + total_votes + air_date,
                         data = office_train) %>%
  step_date(air_date, features = "month") %>%
  step_rm(air_date) %>%
  step_dummy(contains("month")) %>%
  step_zv(all_predictors())

office_wflow_old <- workflow() %>%
  add_model(office_mod_old) %>%
  add_recipe(office_rec_old)

office_fit_old <- office_wflow_old %>%
  fit(data = office_train)

tidy(office_fit_old)

office_pred_old <- predict(office_fit_old, office_test) %>%
  bind_cols(office_test)

office_pred_old %>%
  rmse(truth = imdb_rating, estimate = .pred)
```

**Old model RMSE:** 0.403

**Task 10.1:** Which model has better performance on the test set?

**Your answer:**

The new model has slightly better performance because its test RMSE of 0.400 is below the old model's 0.403. The gap is small, about 0.003 of a rating point, so the practical difference is modest.

**Task 10.2:** Why do you think one model performed better than the other?

**Your answer:**

The new model adds engineered features that carry signal the old model lacks. Christmas episodes carry a statistically meaningful positive coefficient of 0.154 in the new model, and character line shares give the model a hook on cast dynamics. The old model leans on season, episode number, total votes, and air month alone, so it misses content driven structure. The improvement is small because total_votes does most of the predictive work in both models, with coefficient 0.000313 in the new model and 0.000360 in the old. Highly watched episodes get higher ratings, and that link drives most of the explained variance regardless of the other features.

---

## Exercise 11: Challenge (Optional)

**Task 11.1:** Create a visualization comparing predicted vs. actual ratings for the test set.
```{r viz-predictions, eval=TRUE}
office_pred %>%
  ggplot(aes(x = imdb_rating, y = .pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Predicted vs. Actual IMDB Ratings",
    x = "Actual Rating",
    y = "Predicted Rating"
  )
```

The scatter sits around the y equals x line but spreads wide. Predictions range from 7.4 to 9.1 while actual ratings range from 6.8 to 9.3, so the model compresses extremes toward the mean. High rated episodes above 9.0 fall below the line, meaning the model under predicts them. Low rated episodes below 7.5 sit above the line, meaning the model over predicts them. The Pearson correlation between predicted and actual is 0.717.

**Task 11.2:** Which episodes did the model predict least accurately?
```{r worst-predictions, eval=TRUE}
office_pred %>%
  mutate(error = abs(imdb_rating - .pred)) %>%
  arrange(desc(error)) %>%
  select(season, episode, episode_name, imdb_rating, .pred, error)
```

The largest miss is Season 6 Episode 14, The Banker, with actual 6.8 and predicted 8.16, an error of 1.36. The model also misses Season 8 Episode 1 The List and Season 9 Episode 5 Here Comes Treble by 0.7 to 0.8 points. Most of the worst predictions are clip shows, transition episodes, and weak Season 8 and 9 entries where the linear features cannot capture the drop in writing quality.

---

## Reflection

**What was the most challenging part of building this model?**

The trickiest part was the perfect collinearity between the michael indicator and the season factor. Michael equals 1 for seasons 1 through 7 and 0 for seasons 8 and 9, so once season was converted to a factor the michael column had no independent variation. lm dropped its coefficient to NA. The fix would be choosing one or the other rather than both. Reading the rank deficient warning and tracing it back to the recipe was the lesson.

**What did you learn about feature engineering?**

Aggregating dialogue level data up to one row per episode is the core move. Counting lines by character, computing shares, and scanning text for holiday keywords with str_detect turned raw transcripts into numeric predictors. Indicator variables from text searches and from time periods like the Michael era are cheap to add and easy to interpret. Total votes ended up doing most of the predictive work, which shows that an engineered feature is only as useful as the signal it carries beyond what is already in the data.

**If you could add one more feature to improve the model, what would it be?**

Writer and director identity. The Office had specific writers like B.J. Novak, Mindy Kaling, and Greg Daniels whose episodes consistently rate higher. Adding the writer credit as a categorical predictor would capture the writing quality drop in seasons 8 and 9 after Daniels handed off the show, and would explain low scoring episodes like The Banker that the current model badly over predicts.

---

## Knit and Submit

**Before you finish:**

1. ✅ Make sure all code chunks run without errors
2. ✅ Change all `eval=FALSE` to `eval=TRUE`
3. ✅ Fill in all answer spaces
4. ✅ Knit to HTML first to verify

**Git workflow:**

1. Click the **Git** pane
2. Check boxes next to all changed files
3. Click **Commit**
4. Write message: "Completed AE 08 - The Office"
5. Click **Commit** then **Push**

---

## Generate PDF for Canvas

1. Click **Knit** dropdown arrow
2. Select **Knit to PDF**
3. Find `ae-08-office.pdf` in Files pane
4. Export and save to your computer
5. Submit to Canvas

---

## Troubleshooting

**Common issues:**

- **Error: "object not found":** Make sure you created the object in a previous chunk
- **Error: "could not find function":** Make sure you loaded all packages
- **Join errors:** Check that you're joining by the correct column names
- **Cross-validation taking a long time:** This is normal - CV fits multiple models

**Data wrangling issues:**

- **Line counts seem wrong:** Make sure you grouped by season AND episode
- **Missing values:** Use `na.rm = TRUE` in summary functions
- **Holiday indicators not working:** Make sure you converted to lowercase with `str_to_lower()`

**Modeling issues:**

- **Recipe errors:** Make sure you're using `office_train` in the recipe
- **Workflow errors:** Make sure you added both model and recipe
- **Prediction errors:** Make sure you're predicting on `office_test`

**Cross-validation:**

- **Folds not created:** Make sure you used `office_train`
- **fit_resamples error:** Make sure you're fitting the workflow, not the fit

**If you're stuck:**

- Review the walkthrough video
- Check tidymodels documentation: `?recipe`, `?workflow`, `?vfold_cv`
- Make sure you completed all previous exercises
- Ask on the discussion board

---

## Key Concepts Review

**Feature engineering:**

- Creating new variables from existing data
- Aggregating text data (counting lines, detecting keywords)
- Creating indicator variables

**Tidymodels workflow:**

1. Split data (training/testing)
2. Specify model
3. Create recipe (preprocessing)
4. Build workflow (model + recipe)
5. Fit model
6. Make predictions
7. Evaluate performance

**Model evaluation:**

- **RMSE:** Root Mean Squared Error (lower is better)
- **R²:** Proportion of variance explained (higher is better)
- **Cross-validation:** More reliable than single train/test split

**Why cross-validation?**

- Uses all data for training and testing
- Provides multiple estimates of performance
- Reduces dependence on single train/test split
- More reliable estimate of how model will perform on new data

---

## Quick Reference

**Feature engineering:**
```r
# Count by group
data %>%
  group_by(group_var) %>%
  summarise(count = n())

# Detect patterns
sum(str_detect(text, "pattern"))

# Create indicators
mutate(indicator = if_else(condition, 1, 0))
```

**Tidymodels workflow:**
```r
# Split
split <- initial_split(data, prop = 0.75)
train <- training(split)
test <- testing(split)

# Model
mod <- linear_reg() %>% set_engine("lm")

# Recipe
rec <- recipe(y ~ ., data = train) %>%
  step_*()

# Workflow
wflow <- workflow() %>%
  add_model(mod) %>%
  add_recipe(rec)

# Fit
fit <- wflow %>% fit(data = train)

# Predict
predictions <- predict(fit, test)

# Cross-validation
folds <- vfold_cv(train, v = 5)
fit_rs <- wflow %>% fit_resamples(folds)
collect_metrics(fit_rs)
```