---
title: "AE 03 - Visualizing Star Wars Characters"
author: "Kaden Zhu"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Introduction

In this application exercise, you'll practice creating various types of visualizations using the Star Wars characters dataset. This dataset contains information about characters from the Star Wars universe, including their physical characteristics, homeworld, and species.

**In this application exercise, you will:**

- Explore data with `glimpse()`
- Create scatterplots with multiple aesthetics
- Make histograms with appropriate binwidths
- Create visualizations for different variable types
- Interpret relationships between variables
- Customize plot labels and aesthetics
- Choose appropriate plot types for different data types

**Estimated time:** 30-45 minutes

---

## Getting Started

### Fork and Clone Workflow

**Quick reminder:** You should be working in YOUR forked `application-exercises` repository.

1. Navigate to the `ae-03-starwars` folder in YOUR fork
2. Open `ae-03-starwars.Rmd`
3. Update the YAML with your name
4. Click **Knit** to verify it works

**If you need detailed fork-and-clone instructions, see AE 01.**

---

## Packages
```{r load-packages, include=FALSE}
library(tidyverse)
```

**What this package does:**
- **tidyverse:** Includes ggplot2 for visualization and dplyr for data wrangling

---

## The Data

The `starwars` dataset comes built-in with the tidyverse. It contains information on Star Wars characters from the first seven films.

---

## Exercise 1: Glimpse at the Data

Glimpse at the `starwars` data frame to understand its structure.
```{r glimpse-starwars}
glimpse(starwars)
```

**Answer the following questions:**

**How many observations (rows) are there?** 87

**How many variables (columns) are there?** 14

**Name three numerical variables:**

1. height
2. mass
3. birth_year

**Name three categorical variables:**

1. gender
2. species
3. homeworld

**What does each observation represent?**
each observations granularity is one character from starwars so each observation holds the variables for a different character in starwars from the first 7 films.
___________________________________________________________________________

---

## Exercise 2: Understanding Aesthetics

The plot below maps several variables to aesthetics. However, there's an issue with how color is being used.

**Original code:**
```{r scatterplot-original, eval=FALSE}
ggplot(starwars, 
       aes(x = height, y = mass, color = gender, size = birth_year)) +
  geom_point(color = "pink")
```

**Task 2.1:** Run the code above. What happened to the `color = gender` mapping?

**Your answer:**
The color = gender mapping disappears because it is overrided by the the color that is set outside of the aesthetic argument
___________________________________________________________________________
___________________________________________________________________________

**Task 2.2:** Why did this happen?

**Hint:** Think about where `color = "pink"` appears vs. where `color = gender` appears.

**Your answer:**
This happens because aes() sets default mappings and that hierarchically precedes the color pink that's being set as a fixed value outside of aes() basically geom_point() can override them. 
___________________________________________________________________________
___________________________________________________________________________

**The issue:** When you set `color = "pink"` inside `geom_point()`, it overrides the aesthetic mapping `color = gender` from `aes()`.

**Task 2.3:** Modify the code below to keep all the aesthetic mappings AND make the points pink:
```{r scatterplot-pink}
ggplot(starwars, 
       aes(x = height, y = mass, size = birth_year, color = gender)) +
  geom_point() +
  scale_color_manual(values = c("pink", "pink", "pink"))
```

**What this plot shows:**

- x-axis: Height of characters
- y-axis: Mass (weight) of characters  
- Size of points: Birth year (larger = born later)
- All points are pink

---

## Exercise 3: Adding Labels

Add appropriate labels for the title, x-axis, y-axis, and size legend. Uncomment the lines and fill in the blanks.
```{r scatterplot-labels}
ggplot(starwars, 
       aes(x = height, y = mass, color = gender, size = birth_year)) +
  geom_point(color = "#30509C") +
  labs(
    title = "Relationship Between Height and Mass of Star Wars Characters",           # Title about relationship between height and mass
    x = "Height CM",               # What is on the x-axis?
    y = "Mass KG",               # What is on the y-axis?
    size = "Birth Year"             # What does size represent?
  )
```

**Hints:**

- Title should describe what the plot shows
- x and y labels should include units if relevant (height in cm, mass in kg)
- size label describes what the size of points represents

**Task 3.1:** Why is there no `color` label in the `labs()` function?

**Your answer:**
There is no color label in labs() because color is set to a fixed value in geom_point(), so it is not mapped to a variable. Since there is no color legend, there is no need to label it.

___________________________________________________________________________

---

## Exercise 4: Create a Histogram

Pick a single numerical variable and make a histogram of it. Select a reasonable binwidth.

**First, explore the variable:**
```{r explore-for-histogram}
# Choose a variable and look at its range
starwars %>%
  summarise(
    min_height = min(height, na.rm = TRUE),
    max_height = max(height, na.rm = TRUE),
    mean_height = mean(height, na.rm = TRUE)
  )
```

**Now create your histogram:**
```{r histogram, eval = FALSE}
# Fill in the blanks
ggplot(starwars, aes(x = height)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Heights of Characters in Star Wars",
    x = "Height CM",
    y = "Number of Characters"
  )
```

**Hints:**

- Choose from: height, mass, birth_year
- Try different binwidths: 10, 20, 50 (depending on your variable)
- Use `fill` to color the bars, `color` to color the borders

**Task 4.1:** What binwidth did you choose and why?

**Your answer:**

I chose a binwidth of 10 because it shows the distribution clearly without making the bars too wide or too narrow. A smaller binwidth would make the graph look noisy, while a larger binwidth would hide important variation in heights. 
**Task 4.2:** Describe the distribution. Is it symmetric, skewed left, or skewed right? Are there any outliers? 

**Your answer:**
The distribution is slightly right skewed because there are only a few very tall characters that are above 250 cm that extend the right tail. Most characters cluster between about 160 and 200 cm. There are a few short characters below 100 cm, which may also be considered outliers.

___________________________________________________________________________
___________________________________________________________________________

---

## Exercise 5: Numerical vs. Categorical

Pick a numerical variable and a categorical variable and make a visualization to visualize the relationship between the two variables.

**Suggested plot types:**

- Boxplot: `geom_boxplot()`
- Violin plot: `geom_violin()`
- Multiple histograms: `geom_histogram()` with `facet_wrap()`

**Example combinations to try:**

- height by gender
- mass by species
- birth_year by sex
```{r num-cat, eval = FALSE}
# Your code here
ggplot(starwars, aes(x = height, y = species)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Relationship between Height and Species in Star Wars",
    x = "Height CM",
    y = "Species"
  )
```

**Interpretation:**

What does this visualization tell you about the relationship between these two variables?

he visualization shows that height varies significantly across species. Some species, such as Wookiees and Kaminoans, tend to have higher median heights, while others, such as Ewoks and Yoda’s species, are much shorter. Certain species show a wider spread in heights, indicating more variation within that group like droids, while others have very small ranges, suggesting more consistent heights. There are also some outliers present, meaning a few characters are unusually tall or short compared to others in their species such as the droid that is 200 CM.
___________________________________________________________________________
___________________________________________________________________________
___________________________________________________________________________

---

## Exercise 6: Bar Plot of Categorical Variable

Pick a single categorical variable from the dataset and make a bar plot of its distribution.

**Suggested variables:**

- gender
- sex  
- species
- eye_color
- hair_color
```{r barplot,, eval = FALSE}
# Your code here
ggplot(starwars, aes(x = eye_color)) +
  geom_bar(fill = "coral") +
  labs(
    title = "Distribution of Eyecolors of Starwars characters",
    x = "Eye Color",
    y = "Characters"
  )
```

**Task 6.1:** Which category has the most observations?

**Your answer:** Brown Eye Color

**Task 6.2:** If your plot has many categories with small bars that are hard to read, try flipping the coordinates:
```{r barplot-flipped, eval = FALSE}
# Same code as above, but add coord_flip() at the end
ggplot(starwars, aes(x = eye_color)) +
  geom_bar(fill = "coral") +
  labs(
    title = "Distribution of Eyecolors of Starwars characters",
    x = "Eye Color",
    y = "Characters"
  ) +
  coord_flip()
```

---

## Exercise 7: Two Categorical Variables

Pick two categorical variables and make a visualization to visualize the relationship between the two variables.

**Suggested plot types:**

- Stacked bar plot: `geom_bar()`
- Grouped bar plot: `geom_bar(position = "dodge")`
- Faceted bar plot: `geom_bar()` with `facet_wrap()`
- Heatmap: `geom_tile()` with counts

**Suggested combinations:**

- gender and species
- sex and eye_color
- gender and hair_color

**Option 1: Stacked bar plot**
```{r cat-cat-stacked, eval = FALSE}
# Your code here
ggplot(starwars, aes(x = gender, fill = species)) +
  geom_bar() +
  labs(
    title = "Distribution of Species Across Genders in Star Wars",
    x = "Gender",
    y = "Count",
    fill = "Species"
  )
```

**Option 2: Grouped bar plot (side-by-side)**
```{r cat-cat-grouped, eval = FALSE}
# Your code here
ggplot(starwars, aes(x = gender, fill = species)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Species Distribution by Gender in Star Wars",
    x = "Gender",
    y = "Count",
    fill = "Species"
  )
```

**Choose one of the above and interpret:**

What does this visualization tell you about the relationship between these two variables?

The visualization shows how species are distributed across different genders. Most characters identified as male are Human, while fewer species appear frequently among female or non binary/NA categories. Some species are represented by only one or two individuals. Overall, Humans are represented most across all gender categories, suggesting that the dataset is heavily concentrated in that species.

---

## Exercise 8: Multi-Variable Visualization

Pick two numerical variables and two categorical variables and make a visualization that incorporates all of them.

**Suggested approach:**

- x-axis: numerical variable 1
- y-axis: numerical variable 2
- color: categorical variable 1
- facets: categorical variable 2

**Example combination:**

- height (numerical)
- mass (numerical)
- gender (categorical - use for color)
- species (categorical - use for facets, maybe filter to top 3-4 species)

**Starter template:**
```{r multi, eval=FALSE}
# You might want to filter first to avoid too many facets
starwars_filtered <- starwars %>%
  filter(species %in% c("Human", "Droid", "Wookiee", "Gungan"))  # Pick 3-4 species

# Your plot here
ggplot(starwars_filtered, 
       aes(x = height, y = mass, color = gender)) +
  geom_point(size = 3) +
  facet_wrap(~ species) +
  labs(
    title = "Height vs Mass by Species and Gender",
    x = "Height CM",
    y = "Mass KG",
    color = "Gender"
  )
```

**Interpretation:**

What patterns do you notice? How do the relationships differ across the facets?

The plot shows the relationship between height and mass across different species, separated into facets. Overall, height and mass appear positively related, meaning taller characters tend to weigh more. However, the strength of this relationship varies by species. Wookiees are generally taller and heavier than other species. Droids show more variation in mass relative to height, suggesting less consistency in their physical proportions. Humans display a moderate positive relationship with some variation. This indicates that species plays an important role in shaping physical characteristics.

---

## Exercise 9: Reflection

**Task 9.1:** Which type of plot was most useful for exploring this data? Why?

I thought the multi variable scatter plot was the most useful because it allowed me to visualize the relationship between two numerical variables while also comparing patterns across species and gender. It showed how trends differed between groups and made it easier to see both overall patterns and subgroup differences at the same time.
___________________________________________________________________________
___________________________________________________________________________

**Task 9.2:** What was the most interesting pattern or relationship you found?

The most interesting pattern was how much taller and heavier Wookiees are compared to Humans. Creating a boxplot to examine the relationship between species and height clearly showed the variation within each species. When pairing that with the scatter plots, it became clear that not all species follow a consistent height to mass pattern. For example, Droids show more variation and do not follow the same proportional trend as biological species.

___________________________________________________________________________
___________________________________________________________________________

**Task 9.3:** What did you find most challenging about creating these visualizations?

The most challenging part was deciding which variables to combine and how to limit the number of categories so the visualizations remained readable. Some categorical variables had too many levels, which made the plots cluttered and more difficult to interpret without filtering.
___________________________________________________________________________

---

## Knit and Submit

**Before you finish:**

1. ✅ Make sure all code chunks run without errors
2. ✅ Fill in all answer spaces and interpretations
3. ✅ Knit to HTML first to verify everything works
4. ✅ Review the HTML output - check that all plots display correctly

**Git workflow:**

1. Click the **Git** pane
2. Check boxes next to all changed files
3. Click **Commit**
4. Write a commit message: "Completed AE 03 - Star Wars"
5. Click **Commit**
6. Click **Push**
7. Verify your work is on GitHub in **YOUR fork**

---

## Generate PDF for Canvas Submission

Once everything works in HTML:

1. Click the **Knit** dropdown arrow
2. Select **Knit to PDF**
3. Wait for PDF generation
4. Find `ae-03-starwars.pdf` in the Files pane
5. Export and save to your computer
6. Submit the PDF to Canvas

**Remember:** Submit to BOTH GitHub (your .Rmd) AND Canvas (the PDF)!

---

## Troubleshooting

**Common issues:**

- **Error: "object not found":** Make sure you loaded the tidyverse package
- **Plot doesn't show up:** Make sure the code chunk is set to run (not `eval: false`)
- **Too many categories in plot:** Filter the data first or use `coord_flip()` for better readability
- **NAs in plot:** This is normal for this dataset - some characters have missing values

**Aesthetic mapping issues:**

- **Color not showing:** Make sure `color = variable` is inside `aes()`, not `geom_point()`
- **Want to set all points to one color:** Put `color = "colorname"` OUTSIDE `aes()`, inside `geom_point()`

**Plot type selection:**

- **One numerical variable:** histogram or density plot
- **One categorical variable:** bar plot
- **Numerical vs. categorical:** boxplot, violin plot, or faceted histogram
- **Two categorical:** stacked/grouped bar plot or faceted plot
- **Two numerical:** scatterplot
- **Multiple variables:** use color, size, shape, facets

**PDF issues:**

- **LaTeX not found:** Install tinytex: `tinytex::install_tinytex()`
- **Plots too large for PDF:** Adjust `fig.width` and `fig.height` in chunk options

**If you're stuck:**

- Review the walkthrough video
- Check ggplot2 documentation: `?geom_point`, `?geom_histogram`, etc.
- Ask on the discussion board
- Attend office hours

---

## Bonus: Quick Reference

**Basic ggplot structure:**
```r
ggplot(data, aes(x = var1, y = var2)) +
  geom_*() +
  labs(title = "...", x = "...", y = "...")
```

**Common geoms:**

- `geom_point()`: scatterplot
- `geom_histogram()`: histogram
- `geom_bar()`: bar plot
- `geom_boxplot()`: boxplot
- `geom_violin()`: violin plot
- `geom_line()`: line plot

**Common aesthetics (inside `aes()`):**

- `x = variable`: x-axis
- `y = variable`: y-axis
- `color = variable`: color by variable
- `fill = variable`: fill color by variable
- `size = variable`: size by variable
- `shape = variable`: shape by variable

**Fixed aesthetics (outside `aes()`):**

- `color = "blue"`: all points blue
- `fill = "red"`: all filled red
- `size = 3`: all points size 3
- `alpha = 0.5`: transparency (0 = transparent, 1 = opaque)